Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-23 paper.qmd
formatting

These tools represent significant progress in reproducibility, but they share a
common limitation: none addresses system-level dependencies or supports polyglot
environments. A project using [R]{.proglang}, [Python]{.proglang}, and system
tools like [Quarto]{.pkg} or \LaTeX requires coordinating multiple package
managers ([renv]{.pkg} for R, virtual environments for Python, manual
installation for system tools), each with its own configuration and potential
for conflicts.

The most comprehensive approach to reproducibility is containerization with
[Docker]{.pkg}. [Docker]{.pkg} packages a *data product* together with all its
dependencies—including the operating system, system libraries, programming
languages, and packages—into a self-contained image. A statistical analysis can
be viewed as such a data product. The steps to create an image are specified in
a `Dockerfile`, which defines the structure and content of the [Docker]{.pkg}
image.

@@

4. Single-language focus: While [Docker]{.pkg} can support polyglot
  environments, orchestrating them requires manual coordination. There is no
  standard way to specify that one step should run in a Python environment while
  another runs in R.

5. Incomplete Reproducibility: Reproducibility should be understood as a
  spectrum rather than a binary property, a notion emphasised in
  "Reproducibility in Software Engineering" by @dellaieraMasterThesis2024.
  Docker illustrates this spectrum well: it provides strong run-time
  reproducibility—that is, consistency in space (where the same image behaves
  identically across different systems) but weak build-time reproducibility,
  which concerns consistency over time. As defined in the thesis,
  reproducibility at build time means being able to produce an identical build
  artefact regardless of when or where it is built. However, Docker inherently
  fails to guarantee this property. Its reliance on mutable base images and
  non-deterministic instructions such as `apt-get update` leads to temporal
  drift: rebuilding the same `Dockerfile` at different points in time can
  produce distinct outputs. In the experimental evaluation conducted in the
  thesis, even with pinned base images, checksums of the resulting [Docker]{.pkg}
  containers varied across builds and machines. Thus, while Docker offers
  partial reproducibility: strong at run time (space) and weak at build time
  (time), it cannot on its own ensure verifiable, deterministic software
  artifacts. Achieving true reproducibility requires pinning image digests,
  freezing dependency versions, and employing deterministic build systems such
  as Nix or Guix for complete control over both space and time [@Malka_2024].

@@

However, [targets]{.pkg} operates within a single R session. While it can call
Python via [reticulate]{.pkg} or execute system commands, all steps share the
same computational environment. For truly polyglot pipelines (where different
steps may require incompatible dependencies or different language versions) this
limitation becomes problematic. One might run [targets]{.pkg} inside a
[Docker]{.pkg} container to ensure reproducibility, but this requires the entire
pipeline to use a single environment.

@@

- [rix]{.pkg} generates [Nix]{.pkg} expressions from intuitive R function calls,
eliminating the need to learn the [Nix]{.proglang} language. It handles
environment definition including [R]{.proglang}, [Python]{.proglang},
[Julia]{.proglang}, system tools, and their dependencies.

@@

[Nix]{.pkg} is a powerful, cross-platform package manager designed for
installing and building software in a fully reproducible and reliable manner.
Unlike traditional package managers, [Nix]{.pkg} emphasizes immutable
infrastructure and a functional approach, which significantly enhances the
consistency of computational environments. As of writing, the primary Nix
package collection, `nixpkgs`, provides access to over 120,000 packages,
including nearly all of CRAN and Bioconductor. According to Repology, a service
that tracks package repositories across distributions, `nixpkgs` is both the
largest and among the most up-to-date package repositories available (see Figure
@fig-repology).

@@

For projects requiring [Python]{.proglang} packages not yet available in
[Nix]{.pkg}, [rix]{.pkg} supports integration with [uv]{.pkg}, a modern
[Python]{.proglang} package manager. This allows users to combine Nix's
system-level reproducibility (Python interpreter, system libraries) with
[uv]{.pkg}'s comprehensive package availability. While this hybrid approach
sacrifices some of Nix's deterministic build guarantees for Python packages, it
provides a pragmatic solution when the needed packages are unavailable in
`nixpkgs`.

@@

[rixpress]{.pkg} overcomes these limitations by using Nix not just as a package
manager, but as the build automation engine itself. In a [rixpress]{.pkg}
pipeline, each step is defined as a Nix derivation, providing two key benefits:

  1. True Polyglot Pipelines: Each step can have its own Nix environment. A
     Python step can run in a pure Python environment, an [R]{.proglang} step in
     an [R]{.proglang} environment, and a Quarto rendering step in yet another,
     all within the same pipeline.
  2. Deep Reproducibility: Each step is a hermetically sealed Nix derivation
     whose output is cached in the Nix store based on the hash of all its
     inputs. All artifacts are direct children of the computational environment
     (because the computational environment is actually a dependency of the
     artifacts), ensuring they are rebuilt if the environment changes, keeping
     environment and outputs always in sync.

To illustrate these capabilities with a realistic research task, I present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code for this example can be found in the [following
repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

@@

The following example demonstrates how [rixpress]{.pkg} orchestrates a granular,
multi-step data science workflow that crosses language boundaries. (Complete
setup instructions and a detailed walkthrough are provided in Appendix Section
@sec-polyglot-example.)

@@

Computation steps, referred to as *derivations*, are defined using the functions
`rxp_r()`, `rxp_jl()`, and `rxp_py()`, corresponding to [R]{.proglang},
[Julia]{.proglang}, and [Python]{.proglang} environments, respectively. Variants
of these functions suffixed with `_file()` designate *import derivations*, which
introduce external data into the pipeline. It is important to note that any data
imported in this way is stored within the Nix store, ensuring reproducibility
but also persisting the data in the build environment. The function `rxp_qmd()`
is used to compile a [Quarto]{.pkg} document (there is also `rxp_rmd()` which
serves the same purpose for [R Markdown]{.pkg} documents).

@@

Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
pipelines. By leveraging Nix as a build automation engine, [rixpress]{.pkg}
allows polyglot workflows where each step runs in its own hermetically sealed
environment. This ensures deep reproducibility, efficient caching, and seamless
orchestration of polyglot analyses. Most importantly, it provides native
polyglot support without the manual coordination required by containerization
approaches, making complex multi-language workflows accessible to researchers
without systems administration expertise.

@@

The `default.nix` file defines the exact computational environment required to
compile the manuscript, ensuring full reproducibility of the build process. To
guarantee the reproducibility of the manuscript, it is automatically
recompiled via GitHub Actions upon each commit, using the same Nix environment.
An HTML version is additionally deployed to GitHub Pages, providing an
accessible format for viewing on smaller devices at this
[link](https://b-rodrigues.github.io/rix_paper/)^[[https://b-rodrigues.github.io/rix_paper/](https://b-rodrigues.github.io/rix_paper/)].
The PDF version of the manuscript can be found at
[link](https://b-rodrigues.github.io/rix_paper/paper.pdf)^[[https://b-rodrigues.github.io/rix_paper/paper.pdf](https://b-rodrigues.github.io/rix_paper/paper.pdf)].

>>>>

These tools represent significant progress in reproducibility, but they share a
common limitation: none addresses system-level dependencies or supports polyglot
environments. A project using [R]{.proglang} [@r-core-team], [Python]{.proglang}
[@python3], and system tools like [Quarto]{.pkg} or \LaTeX requires coordinating
multiple package managers ([renv]{.pkg} for R, virtual environments for
[Python]{.proglang}, manual installation for system tools), each with its own
configuration and potential for conflicts.

The most comprehensive approach to reproducibility is containerisation with
[Docker]{.pkg}. [Docker]{.pkg} packages a *data product* together with all its
dependencies—including the operating system, system libraries, programming
languages, and packages—into a self-contained image. A statistical analysis can
be viewed as such a data product. The steps to create an image are specified in
a `Dockerfile`, which defines the structure and content of the [Docker]{.pkg}
image.

@@

4. Single-language focus: While [Docker]{.pkg} can support polyglot
  environments, orchestrating them requires manual coordination. There is no
  standard way to specify that one step should run in a [Python]{.pkg}
  environment while another runs in R.

5. Incomplete Reproducibility: Reproducibility should be viewed as a spectrum
   rather than a binary property, as emphasised in *Reproducibility in Software
   Engineering* by @dellaieraMasterThesis2024. [Docker]{.pkg} illustrates this
   spectrum well: it achieves strong run-time reproducibility—consistency across
   systems (space)—but weak build-time reproducibility, which concerns
   consistency across time. As defined in the thesis, build-time reproducibility
   means producing identical artefacts regardless of when or where they are
   built. [Docker]{.pkg} cannot guarantee this, as its reliance on mutable base
   images and non-deterministic commands such as `apt-get update` (or similar
   commands, depending on which base image is used) introduces temporal drift:
   rebuilding the same `Dockerfile` at different times can yield different
   results. Even with pinned base images, the thesis found that container
   checksums varied across builds and machines. Thus, while [Docker]{.pkg}
   provides partial reproducibility, strong in space but weak in time, verifiable,
   deterministic artefacts require pinning image digests, freezing dependencies,
   and using deterministic build systems such as [Nix]{.pkg} for full control
   over both dimensions [@Malka_2024].

@@

However, [targets]{.pkg} operates within a single R session. While it can call
[Python]{.proglang} via [reticulate]{.pkg} or execute system commands, all steps
share the same computational environment. For truly polyglot pipelines (where
different steps may require incompatible dependencies or different language
versions) this limitation becomes problematic. One might run [targets]{.pkg}
inside a [Docker]{.pkg} container to ensure reproducibility, but this requires
the entire pipeline to use a single environment.

@@

- [rix]{.pkg} generates [Nix]{.pkg} expressions from intuitive R function calls,
eliminating the need to learn the [Nix]{.proglang} language. It handles
environment definition including [R]{.proglang}, [Python]{.proglang},
[Julia]{.proglang} [@bezanson2017julia], system tools, and their dependencies.

@@

[Nix]{.pkg} is a powerful, cross-platform package manager designed for
installing and building software in a fully reproducible and reliable manner.
Unlike traditional package managers, [Nix]{.pkg} emphasizes immutable
infrastructure and a functional approach, which significantly enhances the
consistency of computational environments. As of writing, the primary [Nix]{.pkg}
package collection, `nixpkgs`, provides access to over 120,000 packages,
including nearly all of CRAN and Bioconductor. According to Repology, a service
that tracks package repositories across distributions, `nixpkgs` is both the
largest and among the most up-to-date package repositories available (see Figure
@fig-repology).

@@

For projects requiring [Python]{.proglang} packages not yet available in
[Nix]{.pkg}, [rix]{.pkg} supports integration with [uv]{.pkg}, a modern
[Python]{.proglang} package manager. This allows users to combine Nix's
system-level reproducibility ([Python]{.proglang} interpreter, system libraries)
with [uv]{.pkg}'s comprehensive package availability. While this hybrid approach
sacrifices some of Nix's deterministic build guarantees for [Python]{.proglang}
packages, it provides a pragmatic solution when the needed packages are
unavailable in `nixpkgs`.

@@

[rixpress]{.pkg} overcomes these limitations by using [Nix]{.pkg} not just as a
package manager, but as the build automation engine itself. In a
[rixpress]{.pkg} pipeline, each step is defined as a [Nix]{.pkg} derivation,
providing two key benefits:

  1. True Polyglot Pipelines: Each step can have its own [Nix]{.pkg} environment. A
     [Python]{.proglang} step can run in a pure [Python]{.proglang} environment,
     an [R]{.proglang} step in an [R]{.proglang} environment, and a Quarto
     rendering step in yet another, all within the same pipeline.
  2. Deep Reproducibility: Each step is a hermetically sealed [Nix]{.pkg} derivation
     whose output is cached in the [Nix]{.pkg} store based on the hash of all its
     inputs. All artifacts are direct children of the computational environment
     (because the computational environment is actually a dependency of the
     artifacts), ensuring they are rebuilt if the environment changes, keeping
     environment and outputs always in sync.

To illustrate these capabilities with a realistic research task, I present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang}, uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code for this example can be found in the [following
repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

@@

The following example demonstrates how [rixpress]{.pkg} orchestrates a granular,
multi-step data science workflow that crosses language boundaries. (Complete
setup instructions and a detailed walkthrough are provided in Appendix)

@@

Computation steps, referred to as *derivations*, are defined using the functions
`rxp_r()`, `rxp_jl()`, and `rxp_py()`, corresponding to [R]{.proglang},
[Julia]{.proglang}, and [Python]{.proglang} environments, respectively. Variants
of these functions suffixed with `_file()` designate *import derivations*, which
introduce external data into the pipeline. It is important to note that any data
imported in this way is stored within the [Nix]{.pkg} store, ensuring
reproducibility but also persisting the data in the build environment. The
function `rxp_qmd()` is used to compile a [Quarto]{.pkg} document (there is also
`rxp_rmd()` which serves the same purpose for [R Markdown]{.pkg} documents).

@@

Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
pipelines. By leveraging [Nix]{.pkg} as a build automation engine, [rixpress]{.pkg}
allows polyglot workflows where each step runs in its own hermetically sealed
environment. This ensures deep reproducibility, efficient caching, and seamless
orchestration of polyglot analyses. Most importantly, it provides native
polyglot support without the manual coordination required by containerization
approaches, making complex multi-language workflows accessible to researchers
without systems administration expertise.

@@

The `default.nix` file defines the exact computational environment required to
compile the manuscript, ensuring full reproducibility of the build process. To
guarantee the reproducibility of the manuscript, it is automatically recompiled
via GitHub Actions upon each commit, using the same [Nix]{.pkg} environment. An
HTML version is additionally deployed to GitHub Pages, providing an accessible
format for viewing on smaller devices at this
[link](https://b-rodrigues.github.io/rix_paper/)^[[https://b-rodrigues.github.io/rix_paper/](https://b-rodrigues.github.io/rix_paper/)].
The PDF version of the manuscript can be found at
[link](https://b-rodrigues.github.io/rix_paper/paper.pdf)^[[https://b-rodrigues.github.io/rix_paper/paper.pdf](https://b-rodrigues.github.io/rix_paper/paper.pdf)].
